---
title: "Introduction to tidybrookes"
description: >
  Introduction to tidybrookes
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tidybrookes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(tidybrookes)
```

# Importing data from CSV or delimited data

The standard data format exported from by CUH clinical informatics can be 
imported into R using `read_tidybrookes_csv()` or `read_tidybrookes_delim()`,
which are wrappers around `readr:read_csv()` with defaults set to suit the
formats usually used by CUH.

```{r}
adm_file_path <- tidybrookes_example("adm.csv")
adm_raw <- read_tidybrookes_csv(file = adm_file_path, col_types = "adm")
```

The column names in the raw data are inconsistent, making them difficult to
remember. The collection of `*_rename` functions (`adm_rename()`, 
`fsheet_rename()` etc) can rename these columns to more consistent and
standardised names. All other functions in tidybrookes assume that these
column names have been used.
```{r}
adm_raw <- adm_rename(adm_raw)
```

Similar functions are available for each of the clinical tables, e.g. for
flowsheet data:
```{r}
fsheet_file_path <- tidybrookes_example("fsheet.csv")
fsheet_raw <- read_tidybrookes_csv(file = fsheet_file_path,
                                   col_types = "fsheet") %>%
  fsheet_rename()
```

# Defining and extracting clinical fields

A key aspect of `tidybrookes` is a standardised way of defining how clinical 
fields should be extracted and cleaned. The approach taken in `tidybrookes` is
closer to a 'declarative' approach rather than a 'procedural' approach:

- A 'procedural' approach would involve
describing explicit sequence of steps that should be used to extract and clean
each clinical field. This is the usual way that code is written in `R`, but 
it makes it easier to forget steps, and encourages 'copy-paste' style coding,
which can be more error prone.
- A 'declarative' approach instead involves 'describing' each clinical field. 

The tidybrookes approach is to describe each clinical field, along with
a standard set of fields or small functions that describe how the field
should be extracted or cleaned.

Descriptions are created using `*_add` functions, such as `fsheet_add()`.
For example, a flowsheet description for the National Early Warning Score 2
(NEWS2) using:
```{r}
fsheet_def <- list() %>%
  fsheet_add(
    symbol = "news2",
    title = "NEWS2",
    names = c("NEWS2 score"),
    search_pattern = c("news2", "news"),
    search_exclude = c(),
    silently_exclude_na_when = TRUE,
    silently_exclude_when =
      (value_original == " " |
         value_original == "E" # some kind of typo
      ),
    value_as_number_fn =
      case_when(value_original == "3mm" ~ 3,
                TRUE ~ value_as_number),
    expect_after =
      (value_as_number %in% 0:17))
```
Once a list of descriptions has been created, the definitions are then applied
to the raw data to extract and clean the data:
```{r}
fsheet_data <- fsheet_extract(fsheet_raw_example, fsheet_def)
```

# Annotation of admission (`adm`) and admission-discharge-transfer (`adt`) data

Admission data can be annotated using `adm_annotate()`:
```{r}
adm_data_example <- adm_annotate(adm_raw)
```

# Summarising and joining data from separate clinical data tables

We may wish to extract, for example, all NEWS2 measurements during a hospital
visit; or we may wish to extract the _first_ NEWS2 measurement during a
hospital visit. This requires linking admission data and clinical field tables.
This can be done using functions such as `first_during()` and `all_during()`.

For example, to extract the first NEWS2 measurement during each hospital visit:
```{r}
adm_data_example %>%
  first_during(fsheet_data,
             datetime = measurement_datetime,
             during = "during_visit") %>%
  select(person_id,
         visit_id,
         visit_start_datetime,
         visit_end_datetime,
         contains("news2"))
```
